---
title: "Get started with rtprep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rtprep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # dplyr is a Suggests: the vignette reads without it and runs with it
  eval = requireNamespace("dplyr", quietly = TRUE)
)
```

Every response time analysis makes an exclusion decision before the model is
fitted, and most inherit it. The 200 ms and 3 s cutoffs come from the last
paper, the ±2.5 SD criterion from the field. What that decision does to the
parameters is invisible in the data you have, because the contaminants are
not labelled. `rtprep` addresses both halves of that problem. It gives the
screening rules one interface, so that alternatives can be compared instead
of assumed, and it generates data with the contaminants labelled, so that a
pipeline can be tested instead of trusted.

This vignette walks the whole chain once, inside a dplyr pipeline: screen,
filter, aggregate, estimate, and then check the result against the truth.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
```

## The data

`rt_example` holds four participants, two conditions, and 100 trials per
cell. It was generated by `r_contaminated()`, so every trial carries two
columns a real data set cannot: whether it was a contaminant, and which
process produced it.

```{r data}
head(rt_example)

rt_example |>
  count(id, contam_rate, contaminant) |>
  filter(contaminant)
```

The participants contaminate at different rates, from 2% to 15%. That is
deliberate. The question a preprocessing pipeline has to answer is not only
whether it removes contaminants on average, but whether the error it leaves
behind depends on how much each participant contaminated. We come back to
that at the end.

## Screening

A rule is a constructor that holds parameters; `rt_screen()` applies it.
Whatever the rule, the result is one row per trial with the same four
columns: the keep decision, the probability that the trial came from the
decision process, the rule's label, and the reason for a flag.

```{r screen}
screened <- rt_example |>
  mutate(rt_screen(rt, rule_sd(2.5)), .by = c(id, condition))

head(screened)
```

The unnamed `mutate()` splices the four columns in, and `.by` fits the rule
within each participant and condition. The rule could have been anything
else in the roster with no change to the lines around it. Absolute cutoffs,
the median absolute deviation criterion, the recursive criteria of Van Selst
and Jolicoeur (1994), and the contaminant mixture of Ratcliff and Tuerlinckx
(2002) all come back in this shape:

```{r rules}
rule_cutoff(0.18, 3)
rule_mad(2.5)
rule_recursive("modified")
rule_mixture("lognormal")
```

Filtering is one more verb. `rt_keep()` returns the keep column alone and
says once how many trials it dropped, so the exclusion count sits next to
the exclusion rather than being reconstructed later. Pass `.by` to
`rt_keep()` rather than to `filter()`, as a list of the grouping columns,
because inside `filter()` there is no tidy selection and `c(id, condition)`
would concatenate them. The keep vector is the same either way, and the
count then covers the whole data set in one line.

```{r keep}
clean <- rt_example |>
  filter(rt_keep(rt, rule_sd(2.5), .by = list(id, condition)))

nrow(clean)
```

The per-cell diagnostics, which `mutate()` would strip from the attribute,
come from `screen_fits()`:

```{r fits}
rt_example |>
  reframe(screen_fits(rt, rule_sd(2.5)), .by = c(id, condition)) |>
  select(id, condition, n_trials, n_dropped, prop_dropped, lower, upper)
```

## Did the rule remove what you think it removed?

With real data the story ends here. With generated data it does not, and
this is the point of generating any. The rule's decisions can be set against
the truth:

```{r score}
screened |>
  summarise(
    sensitivity = mean(!.keep[contaminant]),
    specificity = mean(.keep[!contaminant]),
    .by = id
  )
```

The ±2.5 SD criterion finds roughly a fifth to a third of each participant's
contaminants and keeps about 98% of the genuine trials. Which contaminants it
finds depends on where they sit relative to the clean response time
distribution, and splitting the hits by process shows that directly:

```{r score-process}
screened |>
  filter(contaminant) |>
  summarise(found = mean(!.keep), n = n(), .by = process)
```

The delayed start-ups sit in the upper tail, where a symmetric criterion
around the mean reaches them. The anticipations sit at the leading edge,
which in a right-skewed distribution is well inside 2.5 standard deviations
of the mean, so the same criterion removes none of them. The informationless
responses overlap the clean core by construction, and a rule that reads only
response times catches almost none of them, here 1 of 20. A rule's hit rate
is a property of the contaminant's location, not of the rule alone.

## Aggregation and estimation

Screening decides which trials survive; aggregation decides what the
survivors are summarised as. `rt_summary()` returns the inputs the
EZ-diffusion equations need, and `ez_ddm()` inverts them into drift, bound,
and non-decision time. Both take vectors and return one row, so they fit the
same grammar:

```{r estimate}
estimates <- clean |>
  reframe(rt_summary(rt, response), .by = c(id, condition)) |>
  mutate(ez_ddm(mean_rt, var_rt, n_upper / n_trials, n_trials))

estimates |>
  select(id, condition, n_trials, drift, bound, ndt)
```

`rt_summary()` has a `method` argument for the robust (median and IQR) and
mixture-based moments, which act on the same trials without removing any;
`?rt_summary` documents them, and the aggregation article on the package
website puts the routes side by side.

## Does the error track the contamination rate?

The participants differ in how much they contaminated, and the data carry
the drift each cell was generated from. Setting the two side by side shows
what a screen leaves behind:

```{r error}
truth <- rt_example |>
  distinct(id, condition, true_drift, contam_rate)

no_screen <- rt_example |>
  reframe(rt_summary(rt, response), .by = c(id, condition)) |>
  mutate(ez_ddm(mean_rt, var_rt, n_upper / n_trials, n_trials)) |>
  select(id, condition, drift_none = drift)

estimates |>
  select(id, condition, drift_sd = drift) |>
  left_join(no_screen, by = c("id", "condition")) |>
  left_join(truth, by = c("id", "condition")) |>
  mutate(
    error_none = drift_none - true_drift,
    error_sd = drift_sd - true_drift
  ) |>
  select(id, condition, contam_rate, true_drift, error_none, error_sd) |>
  arrange(contam_rate, condition)
```

Without preprocessing every cell's drift is underestimated, and the two
largest errors belong to the two participants who contaminated most.
Screening moves every cell, mostly toward the truth, but the heaviest
contaminator's easy condition stays well below it. Four participants with
100 trials per cell cannot separate the rate's effect from sampling error,
and this vignette does not try to. Whether a participant's error tracks
their own contamination rate needs more participants than this vignette
has. `r_contaminated()` generates them with the truth attached, and the
ground-truth article on the package website
(<https://www.gfrischkorn.org/rtprep/articles/ground-truth.html>) runs that
check on data matched to a task of your own.

## Comparing rules

Because the rules share one return shape, comparing them is one call.
`screen_compare()` applies every rule in a list and reports how much each
removed, how often each pair agrees, and how much their excluded sets
overlap.

```{r compare}
cmp <- screen_compare(
  rt_example$rt,
  list(
    cutoff = rule_cutoff(0.18, 3),
    sd = rule_sd(2.5),
    mad = rule_mad(2.5),
    recursive = rule_recursive("modified"),
    mixture = rule_mixture("lognormal")
  ),
  .by = list(rt_example$id, rt_example$condition)
)
cmp
cmp$agreement
```

Agreement and the Jaccard overlap answer different questions, and diverge
exactly where it matters: two rules that each drop 2% of trials and never
the same ones agree on 96% of decisions and overlap not at all.

## Checking the exclusions

One more check applies to real data, where the truth is not available. If
the fast trials a rule removed were guesses, their accuracy should be at
chance. `check_guessing()` tests that with a Beta-Binomial Bayes factor. The
±2.5 SD criterion removed no fast trial at all in this data set, so there is
nothing for it to test there; an absolute cutoff at 350 ms does remove some:

```{r guessing}
rt_example |>
  mutate(rt_screen(rt, rule_cutoff(0.35, 3)), .by = c(id, condition)) |>
  reframe(check_guessing(.keep, rt, response), .by = id) |>
  select(id, n_tested, prop_upper, bf_01, bf_evidence)
```

Four or five tested trials per participant give anecdotal evidence at best,
and that is the honest reading: the check needs a rule that removes fast
trials, and enough of them, before it can say anything. Where `n_tested` is
zero the test is silent, which is itself informative about the rule.

## Reporting what you did

A preprocessing step is part of the analysis, and a reader cannot repeat it
from "outliers were removed". Four things pin it down: which rule and at
which setting, the grouping the criterion was computed within, how much it
removed, and whether error trials went through the screen with the correct
ones. All four are in the objects the code already produced, so none of them
has to be typed from memory.

The rule and its setting print themselves, and the per-cell counts come from
`screen_fits()`:

```{r reporting-fits}
rule <- rule_sd(2.5)
rule

fits <- rt_example |>
  reframe(screen_fits(rt, rule), .by = c(id, condition))

fits |>
  summarise(
    cells = n(),
    trials = sum(n_trials),
    dropped = sum(n_dropped),
    prop = sum(n_dropped) / sum(n_trials),
    lowest_cell = min(prop_dropped),
    highest_cell = max(prop_dropped)
  )
```

Report the range across cells as well as the total. A criterion that removes
`r sprintf("%.1f%%", 100 * sum(fits$n_dropped) / sum(fits$n_trials))` overall
can be removing much more from one participant than another, and that spread
is the thing this package exists to make visible.

The reasons say what the rule actually caught, which is worth checking before
describing it:

```{r reporting-reasons}
rt_example |>
  mutate(rt_screen(rt, rule), .by = c(id, condition)) |>
  count(.rule, .reason)
```

```{r reporting-values, include = FALSE}
n_cells <- nrow(fits)
n_trials <- sum(fits$n_trials)
n_dropped <- sum(fits$n_dropped)
pct <- sprintf("%.1f", 100 * n_dropped / n_trials)
lo <- sprintf("%.0f", 100 * min(fits$prop_dropped))
hi <- sprintf("%.0f", 100 * max(fits$prop_dropped))
```

Which gives a Methods sentence that can be written from the output rather
than around it:

> Response times were screened with a ±2.5 SD criterion around the mean,
> computed separately within each participant and condition
> (`r n_cells` cells). This removed `r n_dropped` of `r n_trials` trials
> (`r pct`%), between `r lo`% and `r hi`% per cell; all exclusions were slow
> trials. Error trials were screened alongside correct ones, and accuracy was
> computed after screening.

That last clause is the one most often left out and the one that most often
changes the answer. The criterion here never looked at `response`, so error
trials passed through it on their response times alone; a rule that does read
accuracy, such as `rule_ewma()` or `rule_mixture(use_accuracy = TRUE)`, needs
saying explicitly, because it makes the screen and the dependent variable
share information.

## Where to go next

`?rules` documents every rule with the reference it implements and the
columns it adds to the fits table, and `?rules_compose` covers `rule_all()`,
`rule_any()` and `rule_then()`, which combine them. No single conventional
rule reaches both ends of the distribution, so combining a spread criterion
with an accuracy control chart is often better than tuning either.
`?rt_summary` covers the robust, trimmed and mixture aggregation routes,
`?adjust_accuracy` the accuracy correction that goes with the mixture route,
and `?r_contaminated` the three contaminant processes and how to match the
generator to a task of your own. `?rule_oracle` removes exactly the labelled
contaminants, which is the ceiling any real rule is read against.

`?rtprep-glossary` defines the terms the rest of the documentation assumes,
and `?extending` gives the contract a rule from another package has to meet.
