---
title: "Aggregation: from surviving trials to parameters"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them
  eval = requireNamespace("dplyr", quietly = TRUE) &&
    requireNamespace("tidyr", quietly = TRUE) &&
    requireNamespace("ggplot2", quietly = TRUE)
)
# CI must fail rather than publish a page whose chunks silently did not run
if (nzchar(Sys.getenv("CI"))) stopifnot(knitr::opts_chunk$get("eval"))
okabe_ito <- c(
  "#0072B2", "#E69F00", "#009E73", "#D55E00",
  "#56B4E9", "#CC79A7", "#F0E442", "#000000"
)
```

Screening decides which trials survive. Aggregation decides what the
survivors are summarised as, and the two fail differently, which is why
`rtprep` keeps them apart. The same trials summarised three ways give three
different sets of summary statistics, and the parameters follow from those.
This article walks through `rt_summary()`'s three methods and its weights,
the accuracy correction that goes with the mixture route, the `ez_ddm()`
inversion and its scaling convention, and then puts four routes side by
side on `rt_example`, whose cells carry the drift they were generated from.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
library(ggplot2)
```

## What `rt_summary()` returns

The EZ-diffusion equations need three things per cell: the mean response
time, its variance, and the proportion correct. `rt_summary()` returns the
first two and the counts behind the third, one row per call, so it fits
inside `reframe()` the way `rt_screen()` fits inside `mutate()`:

```{r summary}
p3_hard <- rt_example |>
  filter(id == "p3", condition == "hard")

rt_summary(p3_hard$rt, p3_hard$response)
```

`contaminant_prop` is `NA` for every method but `"mixture"`, which is the
only one that estimates it. `version = "4par"` summarises the two response
boundaries separately, for the version of the model that lets the starting
point sit off centre:

```{r fourpar}
rt_summary(p3_hard$rt, p3_hard$response, version = "4par")
```

One caveat carries over from `bmm`. A trial whose response is missing
belongs to neither boundary and is left out of `n_upper`, but it still
counts towards `n_trials`, so the ratio understates accuracy when responses
are missing. Drop those trials before aggregating if that matters.

## Robust moments

`method = "robust"` replaces the mean with the median and the variance with
a robust spread: the interquartile range divided by 1.349, squared, or the
median absolute deviation. This is robustness at the level of the statistic,
after Chávez De la Peña et al. (2026): no trial is removed, and a contaminant
in the tail moves the median very little and the interquartile range not at
all.

```{r robust}
moments <- bind_rows(
  simple = rt_summary(p3_hard$rt, p3_hard$response),
  robust_iqr = rt_summary(p3_hard$rt, p3_hard$response, method = "robust"),
  robust_mad = rt_summary(
    p3_hard$rt, p3_hard$response, method = "robust", robust_scale = "mad"
  ),
  .id = "method"
)
moments |>
  select(method, mean_rt, var_rt)
```

```{r robust-values, include = FALSE}
var_of <- function(m) moments$var_rt[moments$method == m]
```

The median sits well below the mean, as it does for any right-skewed
distribution, and the IQR-based robust variance is
`r round(100 * var_of("robust_iqr") / var_of("simple"))`% of the sample
variance. Those are not corrections for contamination; they are different
statistics, and they feed the same equations. The last section puts the
four routes side by side.

## Mixture moments

`method = "mixture"` fits the same contaminant mixture `rule_mixture()`
fits, then reads the mean and variance off the fitted core rather than off
the data. A trial the uniform component owns contributes nothing to the
moments, and the estimated contaminant proportion comes back alongside
them. This is the summary-level version of modelling the contaminant
instead of deleting it:

```{r mixture}
bind_rows(lapply(c("exgaussian", "lognormal", "invgaussian"), function(core) {
  rt_summary(
    p3_hard$rt, p3_hard$response, method = "mixture", distribution = core
  ) |>
    mutate(distribution = core, .before = 1)
})) |>
  select(distribution, mean_rt, var_rt, contaminant_prop)
```

The `...` arguments go to the fit as in `rule_mixture()`, with the same
defaults, including `maxit = 500` where `bmm::ezdm_summary_stats()` uses
100. When the fit fails the moments fall back to the robust ones with a
warning, as in `bmm`. That differs from `rt_screen()`, where the analogous
failure keeps every trial, and the difference is deliberate: a screen that
cannot be evaluated should remove nothing, but a summary still has to
return a number.

```{r mixture-fallback}
rt_summary(p3_hard$rt, p3_hard$response, method = "mixture", maxit = 2)
```

## Weights: the other way of not letting contaminants count

The mixture screen's `.prob` can go into `rt_summary()` as a weight vector
instead of through a threshold. Weighted moments let a doubtful trial
contribute a little, where mixture moments let a trial the uniform
component owns contribute nothing at all. The two are different models of
the same doubt, and the package refuses to combine them:

```{r weights}
screened <- rt_screen(p3_hard$rt, rule_mixture("lognormal"))

rt_summary(p3_hard$rt, p3_hard$response, weights = screened$.prob)

try(rt_summary(
  p3_hard$rt, p3_hard$response, weights = screened$.prob, method = "mixture"
))
```

The weighted variance uses the reliability-weight denominator
$\sum w - \sum w^2 / \sum w$, which reduces to $n - 1$ when the weights are
equal. Frequency weights would use $\sum w - 1$ and are the wrong model,
because `.prob` is a probability, not a count. The same reasoning governs
`min_trials`: with weights, the comparison uses Kish's effective sample size
$(\sum w)^2 / \sum w^2$, so two unit weights among ninety-eight zeros count
as two trials, and the moments come back `NA` rather than as noise:

```{r kish}
two_trials <- c(rep(1, 2), rep(0, 98))
rt_summary(p3_hard$rt, p3_hard$response, weights = two_trials)
```

## Correcting the accuracy counts

Mixture moments take the contaminants out of the response times, but the
accuracy counts still include them. `adjust_accuracy()` removes the
estimated contaminants from the counts on the assumption that they respond
correctly at `guess_rate`. It is stochastic by design: how many trials were
contaminants, and how many of those happened to be correct, are both
binomial draws, and a point estimate would understate that uncertainty.
There is no `set.seed()` inside the package, so the seed is yours:

```{r adjust}
set.seed(2026094)
adjust_accuracy(n_upper = 84, n_trials = 100, contaminant_prop = 0.10)

# NA or zero leaves the counts alone, so the column of a summary table
# drops in as it is
adjust_accuracy(n_upper = 84, n_trials = 100, contaminant_prop = NA)
```

The function is vectorised over its rows and draws each row independently,
so it takes the columns of a summary table directly. For a single row the
two draws are made in the same order as `bmm::adjust_ezdm_accuracy()`, and
the two functions agree draw for draw from the same seed.

## `ez_ddm()` and the scaling constant

`ez_ddm()` is the closed-form inversion of Wagenmakers, van der Maas, and
Grasman (2007): mean, variance, and accuracy in; drift, bound, and
non-decision time out. It is exported so that the whole chain from screening
to parameters runs with `rtprep` alone. The one argument that needs a
sentence is `s`. Wagenmakers et al. fix the diffusion coefficient at 0.1,
`rtprep` at 1, and the choice is a units convention rather than a modelling
one: drift and bound scale linearly with `s`, non-decision time does not.
Their worked example under both conventions:

```{r scaling}
bind_rows(
  wagenmakers = ez_ddm(
    mean_rt = 0.723, var_rt = 0.112, accuracy = 0.802, n_trials = 100, s = 0.1
  ),
  rtprep = ez_ddm(
    mean_rt = 0.723, var_rt = 0.112, accuracy = 0.802, n_trials = 100
  ),
  .id = "convention"
)
```

A drift of 0.1 at `s = 0.1` and a drift of 1 at `s = 1` describe the same
process. Anything compared across packages or papers has to be on one scale
first; `bmm` uses 1 as well.

The equations divide by the logit of accuracy and break down at 0, 0.5, and
1. The published edge correction moves the offending value by
$1 / (2n)$, and `ez_ddm()` applies it silently, because a warning per cell
would bury a simulation run. Which rows it touched comes back as a column,
so that it survives subsetting and the dplyr verbs:

```{r edge}
edges <- ez_ddm(
  mean_rt = 0.5, var_rt = 0.02, accuracy = c(1, 0.5, 0.9), n_trials = 50
)
edges
```

A cell at perfect accuracy gets a drift of `r round(edges$drift[1], 1)`
rather than infinity, and `edge_corrected` is the flag to count in a
results table.

## Four routes on `rt_example`

The routes above are alternatives to trimming, not additions to it. Four
pipelines on `rt_example`, each applied per cell: no preprocessing, the
±2.5 SD trim with simple moments, robust moments on all trials, and mixture
moments on all trials with the accuracy counts adjusted. Each ends in
`ez_ddm()`, and the data carry the drift each cell was generated from:

```{r routes}
summarise_cells <- function(data, ...) {
  data |>
    reframe(rt_summary(rt, response, ...), .by = c(id, condition))
}

routes <- bind_rows(
  none = summarise_cells(rt_example),
  trimmed = rt_example |>
    filter(rt_keep(rt, rule_sd(2.5), .by = list(id, condition))) |>
    summarise_cells(),
  robust = summarise_cells(rt_example, method = "robust"),
  mixture = summarise_cells(
    rt_example, method = "mixture", distribution = "lognormal"
  ),
  .id = "route"
)

set.seed(2026094)
estimates <- routes |>
  mutate(adjust_accuracy(n_upper, n_trials, contaminant_prop)) |>
  mutate(ez_ddm(mean_rt, var_rt, n_upper_adj / n_trials_adj, n_trials_adj)) |>
  left_join(
    distinct(rt_example, id, condition, true_drift, contam_rate),
    by = c("id", "condition")
  ) |>
  mutate(
    route = factor(route, levels = c("none", "trimmed", "robust", "mixture")),
    error = drift - true_drift
  )

estimates |>
  summarise(
    mean_error = mean(error),
    mean_abs_error = mean(abs(error)),
    .by = route
  )
```

```{r routes-figure, fig.alt = "Dot plot of the estimated drift of each participant and condition under four aggregation routes, with the true drift drawn as a black bar in each cell."}
ggplot(estimates, aes(x = condition, y = drift, colour = route)) +
  geom_point(
    aes(y = true_drift), shape = 95, size = 8, colour = "black"
  ) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_colour_manual(values = okabe_ito[c(8, 1, 3, 2)]) +
  facet_wrap(~id, nrow = 1) +
  labs(x = NULL, y = "Estimated drift", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The table is each route's mean signed and mean absolute error against the
generating drift over the eight cells; the figure puts the four estimates
of every cell next to its true drift. Two mechanical differences separate
the mixture route from the other three, and both enter `ez_ddm()` in the
same direction: its variance is read off the fitted core rather than the
sample, and `adjust_accuracy()` raises the proportion correct; a smaller
variance and a higher accuracy both raise the drift.

Eight cells of 100 trials cannot rank four pipelines, and this page does not
try. What they show is that the same trials summarised four ways give four
drift estimates per cell, and the `route` column is the whole difference
between them. [The ground-truth article](ground-truth.html) carries the
same chain through on data matched to a task of your own, with a
perfect-exclusion oracle as the ceiling.

## Differences from `bmm`

`bmm::ezdm_summary_stats()` returns the same columns under the same names,
and the two functions are compared over the full grid of methods, versions,
and cores in `rtprep`'s test suite. Three defaults differ. `rt_summary()`
defaults to `method = "simple"` where `bmm` defaults to `"mixture"`, because
a package about preprocessing choices should not make one of them silently.
The mixture route defaults to `maxit = 500` where `bmm` uses 100. And for
`version = "4par"` with mixture moments, the contaminant bounds are resolved
once on the pooled response times before the split, so that the two halves
are fitted against the same uniform component. The correspondence between
the two packages, function by function, is the subject of
[its own article](rtprep-and-bmm.html).
