---
title: "Comparing rules: what a different choice would have removed"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them
  eval = requireNamespace("dplyr", quietly = TRUE) &&
    requireNamespace("tidyr", quietly = TRUE) &&
    requireNamespace("ggplot2", quietly = TRUE)
)
# CI must fail rather than publish a page whose chunks silently did not run
if (nzchar(Sys.getenv("CI"))) stopifnot(knitr::opts_chunk$get("eval"))
okabe_ito <- c(
  "#0072B2", "#E69F00", "#009E73", "#D55E00",
  "#56B4E9", "#CC79A7", "#F0E442", "#000000"
)
```

The question `rtprep` exists to make askable is what a different
preprocessing choice would have removed. It is normally unanswerable
because every implementation returns a different shape: a trimmed data
frame here, a vector of probabilities there, a set of summary statistics
somewhere else. With one return shape the comparison is one call, and this
article is about that call: what `screen_compare()` returns, which of its
two agreement measures answers which question, how to see what each rule
removed and why, and how to tabulate the roster's settings and realised
drop rates from the result object.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
library(ggplot2)
```

## One call, several tables

`screen_compare()` takes a named list of rules and applies each one with
the same grouping, policy, and threshold. The get-started vignette runs a
five-rule roster; this one adds the accuracy chart, which needs `response`,
and uses the lognormal mixture:

```{r compare}
roster <- list(
  cutoff = rule_cutoff(0.18, 3),
  sd = rule_sd(2.5),
  mad = rule_mad(2.5),
  recursive = rule_recursive("modified"),
  mixture = rule_mixture("lognormal"),
  ewma = rule_ewma()
)

cmp <- screen_compare(
  rt_example$rt, roster,
  response = rt_example$response,
  .by = list(rt_example$id, rt_example$condition)
)
cmp
```

The print method gives the overall drop rate per rule and the pair that
agrees least. Behind it sit three matrices with one row per trial and one
column per rule, and three tables:

```{r names}
names(cmp)
dim(cmp$keep)
```

`keep`, `prob`, and `reason` are the per-trial columns of `rt_screen()`
stacked side by side; `drops` counts what each rule removed per group, with
a column per reason; `agreement` compares every pair; `fits` stacks each
rule's per-group diagnostics with a `.rule` column. `summary(cmp)` prints
the drops and the agreement in full.

## Agreement and Jaccard answer different questions

```{r agreement}
cmp$agreement
```

`agree` is the share of trials on which two rules made the same decision.
`jaccard` is the overlap of the two sets of removed trials: the size of
their intersection over the size of their union. The two diverge exactly
where it matters, and a constructed pair shows the extreme. On one cell,
a cutoff that removes only the fastest 2% and a cutoff that removes only
the slowest 2% never remove the same trial:

```{r divergence}
p3_hard <- rt_example |>
  filter(id == "p3", condition == "hard")

edges <- screen_compare(
  p3_hard$rt,
  list(
    fast_only = rule_cutoff(min = quantile(p3_hard$rt, 0.02)),
    slow_only = rule_cutoff(max = quantile(p3_hard$rt, 0.98))
  )
)
edges$agreement
```

They agree on `r round(100 * edges$agreement$agree)`% of decisions and
overlap not at all. Agreement alone would call them interchangeable, and
in the roster above the same thing happens to every pair that includes the
chart: it agrees with the other rules on
`r round(100 * min(cmp$agreement$agree[cmp$agreement$rule_y == "ewma"]))`%
of decisions or more and shares no removed trial with any of them, because
it cuts from below and the rest cut from above. The other reading is the
recursive criterion against the mixture, which agree on
`r round(100 * cmp$agreement$agree[cmp$agreement$rule_x == "recursive" & cmp$agreement$rule_y == "mixture"], 1)`%
of decisions and overlap at
`r round(cmp$agreement$jaccard[cmp$agreement$rule_x == "recursive" & cmp$agreement$rule_y == "mixture"], 2)`:
two rules from different families that removed almost the same trials.

One more case is deliberate. When neither rule removed anything there is no
overlap to compute, and `jaccard` is `NA` rather than 1, because calling an
empty comparison a perfect match would make it look like agreement:

```{r jaccard-na}
screen_compare(
  p3_hard$rt,
  list(wide = rule_cutoff(0.1, 5), none = rule_none())
)$agreement
```

## What each rule removed, and why

The `drops` table has one row per rule and group, and a column for every
reason any rule gave. Summing over groups shows the composition of each
rule's removals:

```{r reasons}
cmp$drops |>
  summarise(
    across(c(n_dropped, too_fast, too_slow, contaminant), sum),
    .by = .rule
  )
```

The SD and MAD criteria and the recursive criterion removed nothing but
slow trials, and the cutoff removed nothing at all. The chart removed
nothing but fast ones. The mixture's reason is `"contaminant"`, because a
posterior has no side. Those
are three different bets about where the contamination sits, and the table
shows which one a rule made without the truth being consulted. Per group
the drops vary more than the totals suggest; the chart's rate runs from
`r round(100 * min(cmp$drops$prop_dropped[cmp$drops$.rule == "ewma"]))`%
to
`r round(100 * max(cmp$drops$prop_dropped[cmp$drops$.rule == "ewma"]))`%
across the eight cells, which is its lag showing per participant.

## The plot

The plot method draws the drop rates and returns a ggplot when ggplot2 is
installed, with a base-graphics fallback when it is not, so the usual
additions apply:

```{r plot, fig.alt = "Horizontal bar chart of the proportion of trials dropped by each of six screening rules on rt_example, ordered from the absolute cutoff at zero to the MAD criterion at about eleven percent."}
plot(cmp) +
  labs(title = "Proportion of trials each rule removed") +
  theme_minimal()
```

The agreement table draws as a tile plot of the Jaccard overlap, which
makes the two clusters visible at once: the rules that cut from above
overlap with each other and with nothing else.

```{r jaccard-plot, fig.alt = "Tile plot of the Jaccard overlap between every pair of the six rules. The SD, recursive, and mixture rules overlap strongly with each other, the MAD criterion moderately with them, and the cutoff and the EWMA chart overlap with nothing."}
cmp$agreement |>
  mutate(
    rule_x = factor(rule_x, levels = names(roster)),
    rule_y = factor(rule_y, levels = rev(names(roster)))
  ) |>
  ggplot(aes(rule_x, rule_y, fill = jaccard)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(jaccard, 2)), size = 3.5) +
  scale_fill_gradient(low = "grey92", high = okabe_ito[1], limits = c(0, 1)) +
  labs(x = NULL, y = NULL, fill = "Jaccard") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## The threshold is part of the pipeline

`policy` and `threshold` are forwarded to every rule, and for the mixture
they change what is removed. The same lognormal mixture at three thresholds:

```{r thresholds}
vapply(c(0.3, 0.5, 0.7), function(th) {
  sum(!screen_compare(
    rt_example$rt, list(mixture = rule_mixture("lognormal")),
    .by = list(rt_example$id, rt_example$condition), threshold = th
  )$keep)
}, numeric(1)) |>
  setNames(c("0.3", "0.5", "0.7"))
```

The deterministic rules are unaffected, because their `.prob` is 0 or 1 and
every threshold strictly between falls on the same side.

## With ground truth: score every rule at once

Everything above is available on real data. `rt_example` also carries the
truth, and because `keep` is a matrix, scoring every rule is one line for
sensitivity and one for specificity:

```{r truth}
contaminant <- rt_example$contaminant

round(colMeans(!cmp$keep[contaminant, ]), 2)
round(colMeans(cmp$keep[!contaminant, ]), 3)
```

The most aggressive rule finds the most contaminants and pays the most
genuine trials for them; the cutoff finds nothing because nothing in this
fast task reaches its bounds. [The ground-truth article](ground-truth.html)
shows the same scoring on data matched to your own task.

## The roster as a table

The comparison object carries what it was given: `cmp$rules` holds the rule
objects, each with its `label`, and a column mean of `!cmp$keep` is the
realised drop rate per rule. Together they give one row per pipeline:

```{r report}
tibble(
  pipeline = names(cmp$rules),
  setting = vapply(cmp$rules, function(r) r$label, character(1)),
  prop_dropped = colMeans(!cmp$keep)
)
```

Applying each column of `keep` to what comes next repeats the downstream
analysis per pipeline: filter with a column, aggregate, estimate, and
repeat. [The aggregation article](aggregation.html) carries one such table
through to drift estimates, and
[the ground-truth article](ground-truth.html) adds the column a real data
set never has, which is how far each pipeline sits from the truth.
