---
title: "Ground truth: testing a pipeline on data shaped like yours"
---

```{r, include = FALSE}
can_run <- requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("tidyr", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them
  eval = can_run
)
# CI must fail rather than publish a page whose chunks silently did not run
if (nzchar(Sys.getenv("CI"))) stopifnot(can_run)

# The code of this article lives in one file, ground-truth-chunks.R, that
# another document in the repository reads as well, so the two cannot drift
# apart. The call goes through
# do.call() on purpose: pkgdown runs knitr::purl() on the article after
# rendering it, to find its images, and purl evaluates any literal call of
# the chunk reader it sees, even in a comment, on its own and from the
# package root, where the relative path does not resolve. Written this way
# the line is left alone.
article_dir <- Find(
  function(d) file.exists(file.path(d, "ground-truth-chunks.R")),
  c(".", "vignettes/articles")
)
if (is.null(article_dir)) stop("cannot find ground-truth-chunks.R from ", getwd())
do.call(knitr::read_chunk, list(file.path(article_dir, "ground-truth-chunks.R")))
okabe_ito <- c(
  "#0072B2", "#E69F00", "#009E73", "#D55E00",
  "#56B4E9", "#CC79A7", "#F0E442", "#000000"
)
```

Contamination is invisible in the data you have. A trial that came from a
lapse looks like a trial that came from the decision process, only slower,
or faster, or wrong, and every screening rule is a bet about which of those
it can tell apart. Every published evaluation of a rule scored that bet on a
design that was not yours. This article closes the gap the only way it can
be closed: by generating data shaped like your task, contaminating it the
way your design plausibly would, running the pipeline you intend to use, and
scoring it against the truth the generator carried. That costs a few
seconds, and it is the one situation in which a researcher can find out what
their pipeline actually does.

The code on this page lives in one file, `ground-truth-chunks.R`, next to
the article; sourcing it end to end reruns the whole example.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
library(ggplot2)
```

## Three numbers from a pilot

Locate your task by three numbers a pilot gives directly: the mean correct
response time, the proportion correct, and the coefficient of variation of
correct response times. The example uses a moderate task, the three values
in `pilot` below; the get-started vignette's data is faster, and
[the screening-rules article](screening-rules.html) shows what a slow one
does to an absolute cutoff.

```{r gt-pilot}
```

`ez_ddm()` inverts the three into diffusion parameters. At a very large
trial count it is exactly the inversion the generator's own matching uses,
so no internal function is needed:

```{r gt-match}
```

One check to make by eye: the implied non-decision time. The generator's
matching accepts targets whose non-decision time falls between 0.206 and
0.942 s, the range published diffusion fits span, and a target outside it is
a task no diffusion process reaches with plausible parameters. Here it is
`r round(matched$ndt, 2)` s. A clean draw of 10,000 trials confirms that the
parameters reproduce the three observables:

```{r gt-check-match}
```

The match is not available for the racing generator, whose parameters have
no closed-form inversion; the solver that matches it is internal to the
package, and the diffusion generator is the one to use here.

## Contaminate it the way your design would

Which contaminants a design produces is a question about the design, not
about the data. A task with short stimulus onsets, speed pressure, or a
rhythmic pace invites anticipations. A long, low-stakes, unsupervised
session invites informationless responses and late starts. The example
takes the second case, and it lets participants differ in two things: how
much they contaminate, and how able they are.

```{r gt-design}
```

```{r gt-generate}
```

```{r gt-generate-values, include = FALSE}
n_contaminants <- sum(simulated$contaminant)
```

`r n_contaminants` of `r nrow(simulated)` trials are contaminants, at
participant rates from `r round(100 * min(traits$rate), 1)`% to
`r round(100 * max(traits$rate), 1)`%. Every one of them is labelled, and
every participant's true drift is known, which is what the rest of the page
uses.

## What the rules disagree about

With real data the comparison of rules ends at agreement, because nothing
says which rule is right. It is still worth running first, because the
disagreements tell you where the bet is being placed:

```{r gt-compare}
```

```{r gt-agreement}
```

```{r gt-agreement-values, include = FALSE}
drop_rate <- colMeans(!comparison$keep)
mad_ratio <- drop_rate[["mad"]] / drop_rate[c("recursive", "mixture")]
involves <- function(pairs, rule) pairs$rule_x == rule | pairs$rule_y == rule
ewma_jaccard <- comparison$agreement$jaccard[involves(comparison$agreement, "ewma")]
mad_pairs <- comparison$agreement[
  involves(comparison$agreement, "mad") &
    (involves(comparison$agreement, "recursive") |
      involves(comparison$agreement, "mixture")),
]
mad_ratio_text <- if (diff(round(range(mad_ratio), 1)) == 0) {
  as.character(round(min(mad_ratio), 1))
} else {
  paste(round(min(mad_ratio), 1), "to", round(max(mad_ratio), 1))
}
```

Two things stand out before any truth is consulted. The chart removes a set
of trials `r if (all(ewma_jaccard == 0, na.rm = TRUE)) "no other rule touches: its Jaccard overlap with every other rule is zero" else "the other rules barely touch: its Jaccard overlap with every other rule is near zero"`,
because it cuts from below and the others cut from above. And the ±2.5 MAD
criterion removes `r mad_ratio_text` times as many trials as the recursive
criterion or the mixture, while
agreeing with them on more than `r floor(100 * min(mad_pairs$agree))`% of
decisions, which is the case the agreement statistic was built to expose.

## Score the pipeline against the carried truth

The truth turns the same keep matrix into detection rates, overall and by
contaminant process:

```{r gt-score}
```

```{r gt-score-values, include = FALSE}
score_of <- function(rule, column) scored[[column]][scored$rule == rule]
rt_only_anticipations <- scored$anticipations[scored$rule != "ewma"]
```

`r if (all(rt_only_anticipations == 0)) "No rule that reads only response times finds a single anticipation" else "The rules that read only response times find almost no anticipations"`;
the chart finds `r round(100 * score_of("ewma", "anticipations"))`% of them.
Delayed start-ups are found in proportion to how aggressive the rule is,
from `r round(100 * score_of("recursive", "delays"))`% for the modified
recursive criterion to `r round(100 * score_of("mad", "delays"))`% for
±2.5 MAD, and the aggressive rule pays for it in specificity.
Informationless responses are found by almost nothing: every rule's hit rate
on them sits at or near its own false-alarm rate, because those trials have
the decision process's own timing.

The rest of the page carries one rule forward, the modified recursive
criterion, the one [the screening-rules article](screening-rules.html)
shows was built to hold its false-alarm rate across group sizes; swapping it
for another is one column of `comparison$keep`, which the last section does.
On this data it removes
`r round(100 * score_of("recursive", "dropped"), 1)`% of trials and finds
`r round(100 * score_of("recursive", "sensitivity"))`% of the contaminants.

## Was what you removed what you thought?

Two diagnostics apply here. `check_guessing()` asks whether the fast trials
a rule removed were guesses, by testing their accuracy against chance. It
tests only trials that were both removed and fast, so a rule that removes
nothing from below gives it nothing to test:

```{r gt-guessing}
```

```{r gt-guessing-values, include = FALSE}
pooled_guessing <- check_guessing(ewma_keep, simulated$rt, simulated$response)
n_median <- round(median(per_participant$n_tested))
# The largest Bayes factor for guessing that n_median tested trials can
# produce: an exactly even split, every trial counted as fast.
bf_ceiling <- check_guessing(
  keep = rep(FALSE, n_median),
  rt = rep(0.1, n_median),
  response = rep(c(1L, 0L), length.out = n_median),
  threshold_type = "absolute", rt_threshold = 1
)
```

The recursive criterion has no testable trial. The chart does, and pooled
across participants the evidence is `r sub("_", " ", sub("_guessing$", " guessing", pooled_guessing$bf_evidence))`:
the accuracy of what it removed is `r round(pooled_guessing$prop_upper, 2)`,
because the removed set is mostly genuine trials that happened to be fast,
with the anticipations a minority inside it. Per participant the check
tests a median of `r n_median` trials, and the largest Bayes factor for
guessing that `r n_median` trials can produce, on an exactly even split, is
`r round(bf_ceiling$bf_01, 2)` (`r sub("_", " ", sub("_guessing$", " guessing", bf_ceiling$bf_evidence))`).
The pooled call is the one with the trials behind it; the per-participant
calls spread over the categories the count allows:

```{r gt-guessing-per-participant}
table(per_participant$bf_evidence, useNA = "ifany")
```

The leading-edge shift statistic asks, after a cut at the 5th percentile,
how far the surviving minimum moved toward the original 10th percentile.
Near 1 means the removed mass was displaced from the core, which is what
anticipations look like; near 0 means a genuine edge was cut:

```{r gt-leading-edge}
```

```{r gt-leading-edge-values, include = FALSE}
# The clean draw from the matching step, cut into blocks of n_trials, gives
# a reference for what the statistic does on data with no anticipations.
n_slices <- nrow(clean_check) %/% n_trials
clean_slice <- rep(seq_len(n_slices), each = n_trials)
clean_shift <- tapply(
  clean_check$rt[seq_along(clean_slice)], clean_slice, leading_edge_shift
)
```

Across participants the statistic averages
`r round(mean(shift_by_participant), 2)` with a standard deviation of
`r round(sd(shift_by_participant), 2)`, while anticipations make up
`r round(100 * mean(simulated$process == "leading_edge"), 1)`% of the
trials. The clean draw from the matching step gives the reference: cut into
`r n_slices` blocks of `r n_trials` trials, it averages
`r round(mean(clean_shift), 2)` with a standard deviation of
`r round(sd(clean_shift), 2)`. The function returns a value to compare
across participants or against a clean reference like this one; it defines
no cutpoint.

## What the pipeline leaves behind

The question that matters to an individual-differences study is not the
mean error but whether each participant's error tracks their own
contamination rate. Three pipelines, each ending in `ez_ddm()`: nothing
removed, the recursive criterion, and the oracle that removes exactly the
contaminants.

`rule_oracle()` is that last one. It takes the ground-truth vector and
removes exactly the trials it marks, which makes it useless on real data and
the only honest ceiling on generated data. It is a rule like any other, so it
goes through `rt_screen()` and reports in `screen_fits()` the same way, and
swapping it for a real rule changes one argument. Every pipeline below is
read against it.

```{r gt-estimate}
```

```{r gt-figure}
```

```{r gt-figure-print, fig.alt = "Three scatter plots of estimated against true drift for 20 participants, one per pipeline (none, recursive, oracle), with points coloured by each participant's contamination rate and the identity line dashed."}
drift_plot + theme_minimal() + theme(legend.position = "bottom")
```

```{r gt-estimate-values, include = FALSE}
pipeline_stat <- function(p, stat) {
  s <- estimates |>
    summarise(
      mean_error = mean(error),
      rate_error_r = cor(rate, error),
      validity_r = cor(true_drift, drift),
      .by = pipeline
    )
  s[[stat]][s$pipeline == p]
}
```

On this sample the recursive criterion moves the mean error from
`r round(pipeline_stat("none", "mean_error"), 2)` to
`r round(pipeline_stat("recursive", "mean_error"), 2)`. The correlation
between a participant's rate and their error is the harder quantity: it
moves from `r round(pipeline_stat("none", "rate_error_r"), 2)` to
`r round(pipeline_stat("recursive", "rate_error_r"), 2)` where the oracle
gives `r round(pipeline_stat("oracle", "rate_error_r"), 2)`. The scoring
table above says why the rule cannot close that gap: the informationless
responses, which have the decision process's own timing, are found at the
rule's false-alarm rate, and a participant's share of them scales with their
contamination rate. Twenty participants make these correlations noisy.

## Swap the rule and run it again

Everything on this page is a function of one column of `comparison$keep`.
Swapping the modified recursive criterion for the ±2.5 MAD criterion is one
line, and the same summary puts the two next to the no-preprocessing
baseline and the oracle, with the detection columns the truth provides
alongside:

```{r gt-swap}
scored_all <- bind_rows(
  scored,
  score_rule(rep(TRUE, nrow(simulated))) |> mutate(rule = "none"),
  score_rule(oracle_keep) |> mutate(rule = "oracle")
)

bind_rows(estimates, estimate_drift(comparison$keep[, "mad"], "mad")) |>
  summarise(
    mean_error = mean(error),
    rate_error_r = cor(rate, error),
    .by = pipeline
  ) |>
  inner_join(
    select(scored_all, rule, dropped, sensitivity, specificity),
    by = join_by(pipeline == rule)
  ) |>
  mutate(
    pipeline = factor(pipeline, levels = c("none", "recursive", "mad", "oracle")),
    across(where(is.numeric), \(x) round(x, 3))
  ) |>
  arrange(pipeline)
```

This is the table [the comparison article](comparing-rules.html) builds
from real data, with the columns only generated data has. Twenty
participants and one seed produced it, and it is a property of this
generator, this process mix, and this roster. `n_participants`, `n_trials`,
`process_mix`, and the three numbers in `pilot` are what to change to make
it yours; `ground-truth-chunks.R` sourced end to end reruns the page.
