---
title: "Mixture screening: the contaminant mixture and where it fails"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them
  eval = requireNamespace("dplyr", quietly = TRUE) &&
    requireNamespace("tidyr", quietly = TRUE) &&
    requireNamespace("ggplot2", quietly = TRUE)
)
# CI must fail rather than publish a page whose chunks silently did not run
if (nzchar(Sys.getenv("CI"))) stopifnot(knitr::opts_chunk$get("eval"))
okabe_ito <- c(
  "#0072B2", "#E69F00", "#009E73", "#D55E00",
  "#56B4E9", "#CC79A7", "#F0E442", "#000000"
)
```

Every other rule in `rtprep` draws a line and flags what falls outside it.
`rule_mixture()` fits a model instead: a two-component mixture in which one
component is the decision process and the other is a uniform distribution of
contaminants, following Ratcliff and Tuerlinckx (2002). Its output is a
posterior probability per trial rather than a verdict, which is why
`rt_screen()` separates the probability from the keep decision at all. This
article shows what the fit reports, what its arguments do, and then spends
most of its length on the ways it fails, because a model-based screen fails
differently from a cutoff: it can converge cleanly on an answer that removes
nothing.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
library(ggplot2)
```

## The model and the EM

For a response time $t$ the mixture density is

$$
f(t) = (1 - \pi)\, f_{\text{core}}(t \mid \theta) + \pi\, U(t \mid a, b),
$$

where $\pi$ is the contaminant proportion, $f_{\text{core}}$ a parametric
response time distribution with parameters $\theta$, and $U$ the uniform on
the contaminant bounds $a$ and $b$. Expectation maximisation alternates two
steps: compute each trial's responsibility, the posterior probability that
the uniform component produced it, then refit $\pi$ and $\theta$ with those
responsibilities as weights. `.prob` is one minus the final responsibility,
the posterior probability that the trial came from the decision process.

```{r rule}
rule_mixture("lognormal")
```

The fits table carries the whole fit, one row per group. Everything the
sections below discuss is a column of it:

```{r fits}
p3_hard <- rt_example |>
  filter(id == "p3", condition == "hard")

fit <- screen_fits(p3_hard$rt, rule_mixture("lognormal"))
glimpse(fit)
```

`converged` and `iterations` describe the EM, `contaminant_prop` is
$\hat\pi$, `n_fitted` counts the trials inside the bounds, `collapsed` says
whether the fit landed on the zero-contamination solution, and the `par_`
columns are $\hat\theta$ for the chosen core. `p_correct` and
`accuracy_inverted` belong to the accuracy-informed variant and stay `NA`
without it.

## Three cores

The core distribution is the modelling choice, and the same cell fitted three
ways shows how much it moves:

```{r cores}
cores <- c("exgaussian", "lognormal", "invgaussian")

by_core <- bind_rows(lapply(cores, function(core) {
  screen_fits(p3_hard$rt, rule_mixture(core))
}))

by_core |>
  select(distribution, converged, iterations, contaminant_prop, n_dropped)
```

All three converge, and on this cell they agree on the substance: a
contaminant proportion of a few percent and one or two trials removed. The
posteriors behind those decisions differ more than the decisions do. Plotting
`.prob` against the response time, with the generator's truth as colour,
shows where each core places its doubt:

```{r posterior-figure, fig.alt = "Three panels, one per core distribution, plotting each trial's posterior probability of being valid against its response time, coloured by whether the trial was a contaminant. Most trials sit near probability 1; the slowest trials fall away, and the fastest anticipations are barely distinguished from clean trials. Horizontal dashed lines mark thresholds of 0.3, 0.5, and 0.7."}
posteriors <- bind_rows(lapply(cores, function(core) {
  p3_hard |>
    mutate(rt_screen(rt, rule_mixture(core))) |>
    mutate(core = core)
}))

ggplot(posteriors, aes(rt, .prob, colour = contaminant)) +
  geom_hline(yintercept = c(0.3, 0.5, 0.7), linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.8) +
  scale_colour_manual(values = okabe_ito[c(1, 4)]) +
  facet_wrap(~core) +
  labs(x = "Response time (s)", y = "P(valid)", colour = "Contaminant") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r posterior-values, include = FALSE}
slow_contaminant <- posteriors |>
  filter(contaminant, rt > 1.5, rt < 2) |>
  select(core, rt, .prob)
```

The slow tail is where the doubt lives. The uniform component covers the
whole range at low density, so a trial far out in the tail, where the core
density has thinned, tips toward it. The three cores differ in how far out
that point lies. The ex-Gaussian carries an exponential tail of its own and
absorbs slow trials into it: the contaminant at
`r round(slow_contaminant$rt[1], 2)` s keeps a posterior of
`r round(slow_contaminant$.prob[slow_contaminant$core == "exgaussian"], 2)`
under the ex-Gaussian against
`r round(slow_contaminant$.prob[slow_contaminant$core == "lognormal"], 2)`
under the lognormal. The anticipations at the leading edge are a different
story: they sit close enough to the core's own left edge that none of the
three cores separates them from the fastest genuine trials. Which of the
doubtful trials the default policy removes is then a matter of where the
threshold sits, and on this cell a threshold of 0.3, 0.5, or 0.7 removes
`r sum(!rt_screen(p3_hard$rt, rule_mixture("lognormal"), threshold = 0.3)$.keep)`,
`r sum(!rt_screen(p3_hard$rt, rule_mixture("lognormal"), threshold = 0.5)$.keep)`,
and
`r sum(!rt_screen(p3_hard$rt, rule_mixture("lognormal"), threshold = 0.7)$.keep)`
trials with the lognormal core.

## Bounds

The uniform component needs a range. By default it is the group's observed
range buffered outward by half its width on each side, and floored at zero,
because a uniform whose edges sit exactly on data points makes the mixture
barely identifiable:

```{r bounds-default}
fit |>
  select(bound_lower, bound_upper, n_fitted)
range(p3_hard$rt)
```

Numeric bounds are allowed, and they change the model in two ways. Trials
outside the bounds cannot be contaminants, because the uniform component has
zero density there, so they get `.prob` of exactly 1 whatever the core says.
And they leave the fit, which `n_fitted` shows:

```{r bounds-numeric}
narrow <- rt_screen(p3_hard$rt, rule_mixture("lognormal", bound = c(0.3, 1.5)))

attr(narrow, "fits") |>
  select(bound_lower, bound_upper, bound_excludes_fast, bound_excludes_slow,
         n_fitted, contaminant_prop, n_dropped)

narrow |>
  filter(p3_hard$rt > 1.5) |>
  select(.keep, .prob)
```

The warning says the same thing the two `bound_excludes_` columns say: some
observed trials fall outside the bounds, and the rule cannot flag them. That
is the right behaviour for bounds chosen from the design, where the question
is whether anything inside a plausible window is a contaminant, and the
wrong behaviour for bounds guessed too tight. The mixed form
`bound = c(0.1, "max")` fixes one side and leaves the other to the data.

## Convergence and `maxit`

The EM stops when the log-likelihood changes by less than `tol` or after
`maxit` iterations. A fit that hits `maxit` first has not converged, and the
rule then keeps every trial in that group, because a screen that cannot be
evaluated has no grounds to remove data. The fits table records it, and
`rt_screen()` warns once for the whole call rather than once per group:

```{r maxit}
impatient <- screen_fits(
  rt_example$rt, rule_mixture("lognormal", maxit = 3),
  .by = list(rt_example$id, rt_example$condition)
)
impatient |>
  select(.group, converged, iterations, n_dropped)
```

The default is `maxit = 500`, the setting the companion simulation used
throughout. `bmm::flag_contaminant_rts()` uses 100, and at 100 the lognormal
core left a quarter of its fits unconverged on heavy-tailed data in the
package's own measurement. An unconverged fit is not a wrong fit, but it is
a fit that removed nothing, and a comparison between the two packages should
pass `maxit` explicitly on both sides. `bmm` also returns `NA` probabilities
where its fit fails; `rtprep` keeps the trials instead, because an `NA` would
propagate into `.keep`.

## The posterior as a weight

Because `.prob` is a probability rather than a verdict, it does not have to
pass through a threshold at all. `rt_summary()` takes it as a weight vector
and computes weighted moments, so a doubtful trial contributes a little
instead of everything or nothing:

```{r weights}
screened <- rt_screen(p3_hard$rt, rule_mixture("lognormal"))

bind_rows(
  none = rt_summary(p3_hard$rt, p3_hard$response),
  weighted = rt_summary(p3_hard$rt, p3_hard$response, weights = screened$.prob),
  trimmed = rt_summary(p3_hard$rt[screened$.keep], p3_hard$response[screened$.keep]),
  .id = "route"
) |>
  select(route, mean_rt, var_rt, n_trials)
```

```{r weights-values, include = FALSE}
routes <- bind_rows(
  weighted = rt_summary(p3_hard$rt, p3_hard$response, weights = screened$.prob),
  trimmed = rt_summary(p3_hard$rt[screened$.keep], p3_hard$response[screened$.keep]),
  .id = "route"
)
mean_gap_ms <- 1000 * abs(diff(routes$mean_rt))
```

On this cell the weighted and the trimmed means land within
`r if (mean_gap_ms < 1) "a millisecond" else paste(round(mean_gap_ms), "ms")`
of each other and both well away from the untreated one, which is what a
posterior that is nearly 0 or 1 for most trials produces. Where
the posteriors are more spread out the two routes diverge, and
[the aggregation article](aggregation.html) compares them alongside the
third option, mixture
aggregation, which reads the moments off the fitted core and removes
nothing.

## Where it fails

A contaminant proportion of zero is a fixed point of this EM. If the
responsibilities all go to the core, the M-step returns $\hat\pi = 0$, and
the next E-step has no uniform mass to assign. The rule reports that outcome
as `collapsed = TRUE`, and it is the failure mode to look for, because it
arrives with `converged = TRUE` and a plausible log-likelihood. Three
constructions show when it happens and what else goes wrong.

### The core decides whether a fast block is found

The archetypal contaminant is a block of fast guesses well below the leading
edge of the genuine responses. Take a core that is itself ex-Gaussian, the
shape most often used to describe empirical response times, and add 40
guesses uniform between 100 and 200 ms:

```{r fast-block-exg}
set.seed(2026093)
exg_core <- 0.45 + rnorm(400, 0, 0.05) + rexp(400, rate = 1 / 0.15)
block <- runif(40, 0.10, 0.20)
truth <- rep(c(FALSE, TRUE), c(400, 40))

fast_block <- function(rt, truth) {
  bind_rows(lapply(cores, function(core) {
    scr <- rt_screen(rt, rule_mixture(core))
    attr(scr, "fits") |>
      transmute(
        distribution, converged, collapsed,
        contaminant_prop, n_dropped,
        sensitivity = mean(!scr$.keep[truth])
      )
  }))
}

fast_block(c(exg_core, block), truth)
```

```{r fast-block-exg-values, include = FALSE}
exg_result <- fast_block(c(exg_core, block), truth)
```

The lognormal and inverse Gaussian cores find the block whole. They also
overshoot: the fitted proportion is
`r round(exg_result$contaminant_prop[exg_result$distribution == "lognormal"], 2)`
against a true `r round(40 / 440, 2)`, and the
`r exg_result$n_dropped[exg_result$distribution == "lognormal"]` trials
removed are the 40 guesses plus
`r exg_result$n_dropped[exg_result$distribution == "lognormal"] - 40` genuine
responses from the leading edge. The ex-Gaussian core reports
`r round(exg_result$contaminant_prop[exg_result$distribution == "exgaussian"], 2)`
and removes
`r exg_result$n_dropped[exg_result$distribution == "exgaussian"]` trials, of
which `r round(40 * exg_result$sensitivity[exg_result$distribution == "exgaussian"])`
come from the block. It absorbs the guesses by widening its Gaussian
part and driving $\tau$ toward zero, and `bmm`'s EM does the same from the
same starting values, so this is a property of the model rather than of
either implementation. The ex-Gaussian is the default in both packages.

Now the same block on a core produced by the diffusion generator at
`rt_example`'s regime, which has a sharper leading edge and a heavier tail:

```{r fast-block-ddm}
set.seed(2026093)
ddm_core <- r_contaminated(
  400,
  par = list(drift = 1.5, bound = 1.2, ndt = 0.30), rate = 0
)$rt

fast_block(c(ddm_core, block), truth)
```

Every core converges, every core collapses, and the block is untouched. The
difference between the two cores is not the block, which is identical, but
the shape the mixture has to explain around it. With this core the
zero-contamination solution is the attractor for all three, and the fitted
proportion says 0 against a true 9%. The rate matters too, and not
monotonically. The lognormal core on the diffusion core with blocks of 20,
40, 80, and 120 guesses:

```{r fast-block-rate}
set.seed(2026093)
by_rate <- bind_rows(lapply(c(20, 40, 80, 120), function(n_block) {
  rt <- c(ddm_core, runif(n_block, 0.10, 0.20))
  screen_fits(rt, rule_mixture("lognormal")) |>
    transmute(n_block, true_prop = n_block / length(rt),
              contaminant_prop, collapsed, n_dropped)
}))
by_rate
```

The block of 20 is found, at the cost of
`r by_rate$n_dropped[by_rate$n_block == 20] - 20` genuine trials removed
with it; the block of 40 and everything larger is lost entirely. There is
no setting of the core that makes the mixture safe on data of unknown
shape; there is only the check, which is `collapsed` and `contaminant_prop`
read against the rate the design makes plausible.

### Accuracy in the likelihood orders better and removes nothing

`use_accuracy = TRUE` puts accuracy inside the likelihood, on the reasoning
that a fast trial that is correct is less likely to be a guess than a fast
trial that is an error. With $y$ the accuracy indicator, $\gamma$ the chance
rate known from the design, and $p_c$ the estimated accuracy of the decision
process,

$$
f(t, y) = (1 - \pi)\, f_{\text{core}}(t \mid \theta)\, p_c^{\,y} (1 - p_c)^{1 - y}
  + \pi\, U(t \mid a, b)\, \gamma^{\,y} (1 - \gamma)^{1 - y}.
$$

$\gamma$ is fixed, $p_c$ is estimated and reported as `p_correct`. The case
it was built for is contamination that response time cannot see:
informationless responses, which the generator produces as trials with the
decision process's timing and chance accuracy. Ranking quality is read off
the posterior as the sensitivity at 90% specificity, the same statistic the
package's tests use:

```{r overlap}
sensitivity_at <- function(prob, truth, specificity = 0.9) {
  cut <- quantile(prob[!truth], 1 - specificity)
  mean(prob[truth] <= cut)
}

set.seed(2026093)
lapses <- r_contaminated(
  1000,
  par = list(drift = 1.5, bound = 1.2, ndt = 0.30),
  process = "informationless", lapse_prop = 0, rate = 0.15
)

rt_only <- rt_screen(lapses$rt, rule_mixture("lognormal"))
joint <- rt_screen(
  lapses$rt, rule_mixture("lognormal", use_accuracy = TRUE),
  response = lapses$response
)

tibble(
  model = c("RT only", "RT and accuracy"),
  contaminant_prop = c(
    attr(rt_only, "fits")$contaminant_prop,
    attr(joint, "fits")$contaminant_prop
  ),
  p_correct = c(NA, attr(joint, "fits")$p_correct),
  n_dropped = c(sum(!rt_only$.keep), sum(!joint$.keep)),
  ranking_sensitivity = c(
    sensitivity_at(rt_only$.prob, lapses$contaminant),
    sensitivity_at(joint$.prob, lapses$contaminant)
  )
)
```

The accuracy term does what it was meant to do in one respect: the ranking
improves. It does not do the other thing. The fitted proportion stays near
1% against a true 15%, and the number of trials removed does not move.
The accuracy factor favours the core on every correct trial, which widens
the basin of the zero-contamination fixed point; in the package's tests, on
a tighter core, the same construction converges to a proportion below
$10^{-6}$ and removes nothing at all. The better ordering is one the keep
policy never gets to act on.

### Accuracy deflates the evidence when contaminants are accurate

A delayed start-up runs the decision process, only later, so it is usually
correct. Under the joint model every correct contaminant has its contaminant
evidence attenuated by $\gamma / p_c$, and that is not neutrality. The
package's tests show it on a tight ex-Gaussian core with a fixed 250 ms
delay, where the RT-only model is well off the floor and the joint model
loses a large part of its sensitivity. On diffusion-shaped data the
comparison cannot even be made, because every core collapses on delayed
start-ups first:

```{r delays}
set.seed(2026093)
delays <- r_contaminated(
  1000,
  par = list(drift = 1.5, bound = 1.2, ndt = 0.30),
  process = "delay", delay_min = 0.3, delay_max = 0.9, rate = 0.15
)

bind_rows(
  lapply(cores, function(core) screen_fits(delays$rt, rule_mixture(core))),
  screen_fits(
    delays$rt, rule_mixture("lognormal", use_accuracy = TRUE),
    response = delays$response
  )
) |>
  transmute(distribution, use_accuracy = !is.na(p_correct), converged,
            collapsed, contaminant_prop, n_dropped)
```

A uniform shift of 300 to 900 ms placed on 15% of trials is a heavy upper
tail, and the cores absorb it as their own tail rather than as
contamination. The deflation itself is a property of the likelihood, not of
this data set: it applies whenever observed accuracy is well above chance,
which in most response time paradigms is always, so a contaminant
proportion below the RT-only estimate should be read as expected rather
than as evidence of a cleaner data set. Two further things
the variant cannot enforce: it assumes contaminants respond at `chance`,
and `response` must be coded correct/error rather than upper/lower boundary,
which `rtprep` cannot tell apart because both are 0/1. When a fit returns
$p_c$ below chance the component labels have swapped, and
`accuracy_inverted` says so. `check_guessing()` is the diagnostic
counterpart of this variant, testing accuracy after the flagging rather
than during it, and given what is shown above it is the safer of the two
instruments.

## What to report

A mixture screen that is reported as "we removed contaminants with a
mixture model" has reported nothing. Four columns of the fits table make it
a result: `distribution`, `contaminant_prop` against the rate the design
made plausible, `n_dropped` or `prop_dropped`, and the share of groups with
`collapsed = TRUE` or `converged = FALSE`. On `rt_example` that table is
one call:

```{r report}
rt_example |>
  reframe(screen_fits(rt, rule_mixture("lognormal")), .by = c(id, condition)) |>
  summarise(
    groups = n(),
    collapsed = sum(collapsed),
    unconverged = sum(!converged),
    mean_contaminant_prop = mean(contaminant_prop),
    mean_prop_dropped = mean(prop_dropped),
    true_rate = mean(rt_example$contaminant)
  )
```

How often the collapse happens at scale, by core and by contaminant
process, is what the companion simulation measured, and its numbers live in
the simulation article. The rule is in the roster because the idea is sound
and widely used; the columns above are how you find out whether it did
anything on your data.
