---
title: "rtprep and bmm: the correspondence, function by function"
---

```{r, include = FALSE}
can_run <- requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("tidyr", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
has_bmm <- can_run && requireNamespace("bmm", quietly = TRUE)
has_trimr <- can_run && requireNamespace("trimr", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them; the comparisons below also need bmm and trimr, and each
  # such chunk carries its own guard
  eval = can_run
)
# CI must fail rather than publish a page whose chunks silently did not run.
# bmm and trimr are not part of the site's dependencies, so their absence is
# reported on the page instead of failing the build: an omitted comparison
# is labelled, never silently blank.
if (nzchar(Sys.getenv("CI"))) stopifnot(can_run)
```

```{r missing-notice, eval = !has_bmm || !has_trimr, echo = FALSE, results = "asis"}
cat(
  "*This page was built without",
  paste(c("bmm", "trimr")[!c(has_bmm, has_trimr)], collapse = " and "),
  "installed, so the live comparisons that need",
  if (has_bmm == has_trimr) "them" else "it",
  "are not shown. The same comparisons run in the package's test suite on",
  "every check.*"
)
```

`bmm` fits measurement models, among them the EZ-diffusion model and the
full diffusion model, and it ships four helpers for the preprocessing that
comes before a fit. `rtprep` is the package for that preprocessing, and it
reimplements those four rather than depending on `bmm`, so that a package
about preprocessing choices carries no model-fitting dependency and every
rule returns the same shape. Reimplementation has a known cost: nothing
catches a silent divergence from the published algorithm except a test that
compares the two. `rtprep`'s test suite runs those comparisons on every
check, and this page runs the same comparisons live, so what it shows is
demonstrated on the versions named at the bottom rather than asserted.

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
```

## The correspondence

| `bmm` | `rtprep` | What differs |
|---|---|---|
| `flag_contaminant_rts(rt, distribution, contaminant_bound, init_contaminant, max_contaminant, maxit = 100)` | `rt_screen(rt, rule_mixture(distribution, bound, init, max_prop, maxit = 500))` | returns P(contaminant); `.prob` is P(valid). Diagnostics as an attribute; `screen_fits()` as a table. `NA` on a failed fit; every trial kept. |
| `ezdm_summary_stats(rt, response, method = "mixture", ...)` | `rt_summary(rt, response, method = "simple", ...)` | same columns, same names; the default method differs, and `maxit` |
| `adjust_ezdm_accuracy(n_upper, n_trials, contaminant_prop, guess_rate)` | `adjust_accuracy(n_upper, n_trials, contaminant_prop, guess_rate)` | identical, draw for draw, and vectorised over rows |
| `validate_fast_guesses(contam_flag, rt_data, response, ..., guess_prob)` | `check_guessing(keep, rt, response, ..., chance)` | takes the keep flag rather than the contaminant flag; returns a one-row data frame rather than a list |

The comparisons below pass `maxit = 500` on both sides, because that is the
one default that differs and a comparison at two different iteration limits
compares nothing.

## Flagging

`bmm::flag_contaminant_rts()` returns the posterior probability that each
trial is a contaminant. `rt_screen()` with `rule_mixture()` returns the
complement, the probability that the trial came from the decision process,
because the same column has to hold the 0 and 1 of a deterministic rule and
"1 means keep" is the convention that makes `filter(.keep)` read correctly.
The mapping is one minus the other:

```{r flag, eval = has_bmm}
p3_hard <- rt_example |>
  filter(id == "p3", condition == "hard")

theirs <- bmm::flag_contaminant_rts(
  p3_hard$rt, distribution = "lognormal", maxit = 500
)
ours <- rt_screen(p3_hard$rt, rule_mixture("lognormal", maxit = 500))

max(abs((1 - ours$.prob) - as.numeric(theirs)))
all(ours$.keep == (as.numeric(theirs) < 0.5))
```

The two agree to the optimiser's tolerance rather than exactly, because
`rtprep` uses the closed-form weighted maximiser for the lognormal and
inverse-Gaussian cores where `bmm` optimises numerically; the test suite
allows the decisions to differ only for a trial whose posterior sits within
0.01 of the cut. The fit diagnostics are the same quantities in two shapes:
`bmm` attaches a one-row data frame with a list column of parameters,
`rtprep` returns one row per group with the parameters as `par_` columns,
so that a comparison across participants is a table without unnesting:

```{r flag-diagnostics, eval = has_bmm}
attr(theirs, "diagnostics") |>
  select(contaminant_prop, converged, iterations, loglik)

screen_fits(p3_hard$rt, rule_mixture("lognormal", maxit = 500)) |>
  select(contaminant_prop, converged, iterations, loglik, par_mu, par_sigma)
```

When the EM does not converge `bmm` returns `NA` probabilities. `rtprep`
keeps every trial in that group and records `converged = FALSE`, because an
`NA` would propagate into `.keep` and a screen that cannot be evaluated has
no grounds to remove data.

## Summary statistics

`bmm::ezdm_summary_stats()` and `rt_summary()` return the same columns
under the same names, for every method and both versions, so a summary
table built with one drops into a workflow written for the other:

```{r summary, eval = has_bmm}
bind_rows(lapply(c("simple", "robust", "mixture"), function(method) {
  ours <- rt_summary(
    p3_hard$rt, p3_hard$response,
    method = method, distribution = "lognormal", maxit = 500
  )
  theirs <- bmm::ezdm_summary_stats(
    p3_hard$rt, p3_hard$response,
    method = method, distribution = "lognormal", maxit = 500
  )
  tibble(
    method = method,
    same_names = identical(names(ours), names(theirs)),
    max_abs_diff = max(abs(unlist(ours) - unlist(theirs)), na.rm = TRUE)
  )
}))
```

The simple and robust moments agree exactly; the mixture moments agree to
the optimiser's tolerance. Two defaults differ. `bmm` defaults to
`method = "mixture"` and `rtprep` to `"simple"`, because a package about
preprocessing choices should not make one of them silently. And `rtprep`
defaults to `maxit = 500` where `bmm` keeps 100;
[the mixture article](mixture-screening.html) shows what the cap does to a
fit that has not converged. For `version = "4par"` with mixture moments both
packages resolve
the contaminant bounds once on the pooled response times before splitting by
boundary, so the two halves are fitted against the same uniform component.

## Accuracy adjustment

Both functions draw the number of contaminants and the number that happened
to be correct as binomials, and they make the two draws in the same order,
so from the same seed they return the same counts:

```{r adjust, eval = has_bmm}
set.seed(2026097)
adjust_accuracy(n_upper = 84, n_trials = 100, contaminant_prop = 0.10)

set.seed(2026097)
bmm::adjust_ezdm_accuracy(n_upper = 84, n_trials = 100, contaminant_prop = 0.10)
```

`rtprep`'s version is vectorised over its rows, so the columns of a summary
table go in directly; `bmm`'s takes one row at a time.

## The guessing check

`check_guessing()` is a port of `bmm::validate_fast_guesses()` with two
interface changes. It takes the *keep* flag, so the `.keep` column of a
screen drops in without negation, and it returns a one-row data frame,
because in practice the result goes into a results table. The argument
`chance` is `bmm`'s `guess_prob`:

```{r guessing, eval = has_bmm}
screened <- rt_screen(
  rt_example$rt, rule_cutoff(0.35, 3),
  .by = list(rt_example$id, rt_example$condition)
)

ours <- check_guessing(screened$.keep, rt_example$rt, rt_example$response)
theirs <- bmm::validate_fast_guesses(
  contam_flag = !screened$.keep, rt_data = rt_example$rt,
  response = rt_example$response
)

c(ours = ours$bf_01, bmm = theirs$bf_01)
c(ours = ours$n_tested, bmm = theirs$n_tested)
```

## The differences, listed

Beyond the two defaults, the packages part in six places, each deliberate:

* **`.prob` is P(valid)**, the complement of what `flag_contaminant_rts()`
  returns.
* **A rule that cannot be evaluated removes nothing.** Fewer than two trials
  in a group, zero spread, a sample below the smallest tabled size for the
  recursive criteria, or an unconverged mixture: `rtprep` keeps the trials and
  says so in the fits table, where `bmm` returns `NA`.
* **Numeric `response` must be coded 0/1.** `bmm` coerces with
  `as.logical()`, which reads a 1 = error / 2 = correct column as perfect
  accuracy; `rtprep` refuses it instead.
* **Warnings come once per call**, naming how many groups failed, rather than
  once per group.
* **Bounds are inclusive.** A trial is flagged only when it falls strictly
  outside `rule_cutoff()`'s bounds; `trimr` uses strict comparisons, so the
  two can disagree on a trial sitting exactly on a bound.
* **`ez_ddm()` uses `s = 1`**, as `bmm` does, where Wagenmakers et al. use
  0.1; drift and bound scale with `s` and non-decision time does not.

## Handing a screened data set to `bmm`

The point of matching the column names is that a `bmm` fit takes `rtprep`'s
output without translation. Both models need Stan, so the calls are shown
and not run. The EZ-diffusion model fits a summary table, which `rt_summary()`
and `adjust_accuracy()` produce per cell:

```{r ezdm-handoff, eval = FALSE}
library(bmm)

ezdm_data <- trial_data |>
  reframe(
    rt_summary(rt, response, method = "mixture", distribution = "lognormal"),
    .by = c(subject, condition)
  ) |>
  mutate(adjust_accuracy(n_upper, n_trials, contaminant_prop))

fit_ez <- bmm(
  formula = bmf(drift ~ condition, bound ~ 1, ndt ~ 1),
  data = ezdm_data,
  model = ezdm(
    mean_rt = "mean_rt", var_rt = "var_rt",
    n_upper = "n_upper_adj", n_trials = "n_trials_adj"
  )
)
```

The full diffusion model fits trial-level data, which a screen filters in
one verb:

```{r ddm-handoff, eval = FALSE}
clean_data <- trial_data |>
  filter(rt_keep(rt, rule_recursive("modified"), .by = list(subject, condition)))

fit_ddm <- bmm(
  formula = bmf(drift ~ condition, bound ~ 1, ndt ~ 1),
  data = clean_data,
  model = ddm(resp1 = "rt", resp2 = "response")
)
```

The alternative to filtering is to keep every trial and hand the mixture's
posterior to the aggregation as a weight; the aggregation article compares
the two.

## `trimr`

`trimr` implements the absolute cutoff, the SD criterion, and van Selst and
Jolicoeur's criteria, and returns trimmed data or per-cell means rather than
per-trial decisions. Its `sdTrim()` trims the slow side and leaves the fast
side to a separate `minRT` argument, so the comparison is on the upper tail;
`rtprep` flags both tails from one rule. The two keep the same trials:

```{r trimr-sd, eval = has_trimr}
trimr_data <- rt_example |>
  transmute(participant = id, condition, rt, accuracy = response)

kept_by_trimr <- trimr::sdTrim(
  trimr_data,
  minRT = 0, sd = 2.5, perCondition = TRUE, perParticipant = TRUE,
  omitErrors = FALSE, returnType = "raw"
)

kept_by_rtprep <- rt_example |>
  filter(rt_keep(rt, rule_sd(2.5), .by = list(id, condition)))

c(trimr = nrow(kept_by_trimr), rtprep = nrow(kept_by_rtprep))
setequal(kept_by_trimr$rt, kept_by_rtprep$rt)
```

The recursive criteria compare on the per-cell means that `trimr` returns:

```{r trimr-recursive, eval = has_trimr}
trimr_means <- trimr::modifiedRecursive(
  trimr_data,
  minRT = 0, omitErrors = FALSE, returnType = "mean", digits = 10
)

rtprep_means <- rt_example |>
  filter(rt_keep(rt, rule_recursive("modified"), .by = list(id, condition))) |>
  summarise(mean_rt = mean(rt), .by = c(id, condition)) |>
  tidyr::pivot_wider(names_from = condition, values_from = mean_rt)

all.equal(
  as.numeric(as.matrix(trimr_means[, c("hard", "easy")])),
  as.numeric(as.matrix(rtprep_means[, c("hard", "easy")]))
)
```

Two documented divergences remain. `trimr` compares strictly and `rtprep`
inclusively, so a trial sitting exactly on a bound is removed by one and kept
by the other. And `trimr`'s modified recursive procedure removes an extreme
by value, which would take every tied copy at once, where `rtprep` removes
it by position; with continuous response times the two agree, as they do
above. `trimr`'s hybrid statistic, the mean of the two condition means, is
recoverable from `rtprep` by summarising the two variants separately, as
`?rule_recursive` shows, and not from any single keep vector.

## Versions this page was built against

```{r versions}
c(
  rtprep = as.character(packageVersion("rtprep")),
  bmm = if (has_bmm) as.character(packageVersion("bmm")) else "not installed",
  trimr = if (has_trimr) as.character(packageVersion("trimr")) else "not installed"
)
```

A release of `bmm` or `trimr` that changed an algorithm would show here,
and in `rtprep`'s check workflow, which installs both and runs the same
comparisons with tighter tolerances than a page can show.
