---
title: "Screening rules: what each one removes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  # dplyr, tidyr, and ggplot2 are Suggests: the article reads without them and
  # runs with them
  eval = requireNamespace("dplyr", quietly = TRUE) &&
    requireNamespace("tidyr", quietly = TRUE) &&
    requireNamespace("ggplot2", quietly = TRUE)
)
# CI must fail rather than publish a page whose chunks silently did not run
if (nzchar(Sys.getenv("CI"))) stopifnot(knitr::opts_chunk$get("eval"))
okabe_ito <- c(
  "#0072B2", "#E69F00", "#009E73", "#D55E00",
  "#56B4E9", "#CC79A7", "#F0E442", "#000000"
)
```

The get-started vignette applies one rule inside a dplyr pipeline and moves
on. This article stays with the rules. It takes each family in `?rules` in
turn and asks the same three questions of it: what it removes, what it
reports about the removal, and where it fails. Everything runs on
`rt_example`, plus two generated data sets where a failure needs a task the
example does not provide. The contaminant mixture gets one paragraph here and
[its own article](mixture-screening.html).

```{r setup, message = FALSE}
library(rtprep)
library(dplyr)
library(ggplot2)
```

## One return shape

A rule is a constructor that holds parameters and does nothing to data.
`rt_screen()` applies it and returns one row per trial with the same four
columns whichever rule it was given. A short vector with a missing value, a
slow trial, and a fast one shows all the values the last column can take:

```{r shape}
rt <- c(0.31, NA, 0.42, 3.20, 0.09)
rt_screen(rt, rule_cutoff(0.18, 3))
```

`.keep` is the decision, `.prob` the probability that the trial came from the
decision process, `.rule` the label, and `.reason` why the rule flagged it. A
missing response time is excluded from every rule's computation and comes
back with `.keep = FALSE`, `.prob = NA`, and `"missing"` as its reason. A kept
trial never carries a reason.

What a rule computed on the way to those decisions sits in a second table,
one row per group of `.by`, and `screen_fits()` returns it directly. Each
family adds its own columns to the four that are always there (`.group`,
`n_trials`, `n_dropped`, `prop_dropped`), and the rest of this article reads
the rules through that table.

## Absolute cutoffs

The oldest screen and still the most common: discard anything faster than a
plausible minimum or slower than a plausible maximum. The pair most often
inherited from the last paper is 180 ms and 3 s.

```{r cutoff}
rt_example |>
  reframe(screen_fits(rt, rule_cutoff(0.18, 3)), .by = c(id, condition))
```

The fits table reports the bounds and nothing else, because there is nothing
else: the rule estimates nothing. On `rt_example` it removes no trial at all.
The fastest response in the data set is
`r round(1000 * min(rt_example$rt))` ms and the slowest
`r round(max(rt_example$rt), 2)` s, so neither bound is ever reached.

Bounds are inclusive. A trial is flagged only when it falls strictly outside
them, so a response time of exactly 180 ms survives `rule_cutoff(0.18, 3)`.
`trimr` uses strict comparisons and would remove it, which is the one place
the two packages disagree on this rule.

```{r inclusive}
rt_screen(c(0.1799, 0.1800, 0.1801), rule_cutoff(0.18, 3))
```

The failure of an absolute cutoff is a failure of transfer. The pair was
written for a task with a particular speed, and it meets a different one.
`rt_example` is a fast task; the companion simulation's slowest anchor cell
has a mean correct response time of 2.5 s, an accuracy of .80, and a
coefficient of variation of .50. `ez_ddm()` inverts those three observables
into diffusion parameters when it is given a very large trial count, and
`r_contaminated()` with `rate = 0` then generates clean data at exactly that
regime:

```{r slow-task}
slow_par <- ez_ddm(
  mean_rt = 2.5, var_rt = 1.25^2, accuracy = 0.80, n_trials = 1e6
)
slow_par

set.seed(2026092)
slow <- r_contaminated(
  2000,
  par = list(drift = slow_par$drift, bound = slow_par$bound, ndt = slow_par$ndt),
  rate = 0
)
c(
  mean_rt = mean(slow$rt),
  cv = sd(slow$rt) / mean(slow$rt),
  accuracy = mean(slow$response)
)
```

```{r slow-cutoff}
slow_fit <- screen_fits(slow$rt, rule_cutoff(0.18, 3))
slow_fit
```

Not one of these trials is a contaminant, and the 180 ms / 3 s pair removes
`r round(100 * slow_fit$prop_dropped)`% of them. The warning above comes from
the package's units check, which fires whenever a response time exceeds 10 s
because that usually means milliseconds were passed as seconds. Here it means
what it says: a slow, variable task produces a handful of genuine responses
that long, and they are the trials a fixed upper bound removes first.

## Location and spread

`rule_sd()` flags trials further than a multiple of a spread statistic from a
centre, both recomputed within each group. With the mean and the standard
deviation it is the ±2.5 SD criterion, modal practice in the field; with the
median and the median absolute deviation it is the criterion Leys et al.
(2013) recommend instead, and `rule_mad()` is the shorthand for that setting.
One constructor covers both because they are one algorithm with a different
location and spread pair, and the fits table shows the pair it used:

```{r spread}
p3_hard <- rt_example |>
  filter(id == "p3", condition == "hard")

bind_rows(
  screen_fits(p3_hard$rt, rule_sd(2.5)),
  screen_fits(p3_hard$rt, rule_mad(2.5))
)
```

Two things follow from the numbers. The median absolute deviation is smaller
than the standard deviation on a right-skewed distribution, so the same
multiplier draws a narrower band and removes more trials, here
`r screen_fits(p3_hard$rt, rule_mad(2.5))$n_dropped` against
`r screen_fits(p3_hard$rt, rule_sd(2.5))$n_dropped`. And the lower bound of
the SD criterion is negative. No trial can be too fast for it: the mean minus
2.5 standard deviations falls below zero on any response time distribution
with the usual skew, which is why a symmetric criterion around the mean never
reaches the leading edge where anticipations sit.

Miller (1991) is the standard warning about the SD criterion. The proportion
of a skewed distribution that falls outside a fixed multiplier depends on the
sample size, so the criterion changes what it removes as trial counts vary.
Measuring that takes clean data at several group sizes. The generator gives
24,000 trials at `rt_example`'s regime, and `.by` splits them into groups of
10 to 200:

```{r sample-size}
set.seed(2026092)
clean <- r_contaminated(
  24000,
  par = list(drift = 1.5, bound = 1.2, ndt = 0.30),
  rate = 0
)

clean_drop <- function(rule, sizes = c(10, 20, 50, 100, 200)) {
  bind_rows(lapply(sizes, function(k) {
    group <- rep(seq_len(nrow(clean) / k), each = k)
    fits <- screen_fits(clean$rt, rule, response = clean$response, .by = group)
    data.frame(
      rule = rule$label, n_trials = k,
      prop_dropped = mean(fits$prop_dropped)
    )
  }))
}

drop_by_n <- bind_rows(
  clean_drop(rule_sd(2.5)),
  clean_drop(rule_recursive("modified")),
  clean_drop(rule_ewma())
)

drop_by_n |>
  tidyr::pivot_wider(names_from = rule, values_from = prop_dropped)
```

```{r sample-size-values, include = FALSE}
drop_at <- function(label, n) {
  100 * drop_by_n$prop_dropped[drop_by_n$rule == label & drop_by_n$n_trials == n]
}
```

Every trial in this table is genuine, so every entry is a false-alarm rate.
The SD criterion removes `r round(drop_at("sd(2.5, mean, sd)", 10), 1)`% of
clean trials in groups of 10,
`r round(drop_at("sd(2.5, mean, sd)", 20), 1)`% in groups of 20, and
`r round(drop_at("sd(2.5, mean, sd)", 200), 1)`% in groups of 200. Small
groups underestimate the spread of a skewed distribution and the multiplier
bites less, which is the dependence Miller described; at the trial counts
most designs use, the rate has settled. The modified recursive criterion in
the second column was built to remove that dependence, and its rate moves
from `r round(drop_at("recursive(modified)", 10), 1)`% at 10 trials to
`r round(drop_at("recursive(modified)", 200), 1)`% at 200. The third column
is a different rule with a different problem, and the next two sections take
them in turn.

## Recursive and moving criteria

Van Selst and Jolicoeur (1994) answered Miller's sample-size problem by
making the multiplier itself depend on the number of trials. `rtprep` ships
their published table, interpolated between the tabulated sizes as the
table's note instructs and as `trimr` does. Three variants share the
constructor:

```{r recursive}
p4_easy <- rt_example |>
  filter(id == "p4", condition == "easy")

bind_rows(
  screen_fits(p4_easy$rt, rule_recursive("moving")),
  screen_fits(p4_easy$rt, rule_recursive("modified")),
  screen_fits(p4_easy$rt, rule_recursive("hybrid"))
)
```

The moving criterion is one pass with the multiplier read off the table for
the group's trial count, so `iterations` is always 1. The modified recursive
procedure sets the largest remaining response time aside while the mean and
standard deviation are computed, removes the most extreme trial at each end
if it falls outside the resulting bounds, and repeats until nothing is
removed; the `criterion` it reports is the one used on the last pass, and
`iterations` counts the passes. The hybrid reports both sets of bounds and
`n_disagree`, the number of trials on which the two variants differ.

That disagreement is what the hybrid's `.prob` encodes. Per trial it is the
mean of the two decisions, so it takes the values 0, 0.5, and 1:

```{r hybrid}
hybrid <- rt_screen(p4_easy$rt, rule_recursive("hybrid"))
table(prob = hybrid$.prob, keep = hybrid$.keep)
```

Under the default policy a trial is kept when `.prob` exceeds 0.5, so a trial
the two variants disagree on is removed: the hybrid keeps only what both keep.
Van Selst and Jolicoeur's published hybrid statistic is something else, the
mean of the two condition means, and no single keep vector reproduces it
because the two means have different denominators. To recover the published
number, summarise the two variants separately and average the results; the
code is in `?rule_recursive`.

## The accuracy control chart

Every rule so far reads response times alone. `rule_ewma()` is the
exponentially weighted moving average cutoff that shipped with DMAT
(Vandekerckhove & Tuerlinckx, 2007), and the only published screen that reads
accuracy. Trials are ordered from fastest to slowest, an exponentially
weighted average of accuracy is accumulated from chance upwards, and the
cutoff is the response time at which that average first clears a control
limit. Below it, accuracy is indistinguishable from guessing. The rule needs
`response`:

```{r ewma}
screen_fits(p3_hard$rt, rule_ewma(), response = p3_hard$response)
```

```{r ewma-truth}
ewma_removed <- p3_hard |>
  mutate(rt_screen(rt, rule_ewma(), response = response)) |>
  filter(!.keep) |>
  count(contaminant, process)
ewma_removed
```

The chart lags, and the lag is a cost rather than an implementation detail.
The average needs a run of correct trials before it clears the limit, so the
cutoff lands past the point where accuracy actually rose, and the genuine
trials just above the guesses go with them. In this cell the cutoff falls at
`r round(1000 * screen_fits(p3_hard$rt, rule_ewma(), response = p3_hard$response)$cutoff_rt)` ms
and removes `r sum(ewma_removed$n)` trials:
`r sum(ewma_removed$n[ewma_removed$contaminant])` anticipations and
`r sum(ewma_removed$n[!ewma_removed$contaminant])` clean responses that
happened to be fast. The sample-size table above shows the same lag
from the other side: with 10 trials per group the chart has barely started
when the group ends, and it removes
`r round(drop_at("ewma(0.01, 1.5)", 10))`% of clean trials; with 200 it
removes `r round(drop_at("ewma(0.01, 1.5)", 200), 1)`%. Ratcliff and Kang
(2021) note that the rule has not found much use. It is here as the published
accuracy-based comparator, and because it is the only rule in the roster that
can see an anticipation at all.

## The mixture, in one paragraph

`rule_mixture()` fits a two-component mixture by expectation maximisation: a
parametric core for the decision process and a uniform component for the
contaminants, following Ratcliff and Tuerlinckx (2002). It is the one rule
whose `.prob` is a genuine posterior rather than 0 or 1, which is why the
keep policy exists as a separate step. Which core to choose, what the bounds
do, what an unconverged fit does, and the three ways the rule fails are the
subject of [the mixture article](mixture-screening.html).

## Probability before decision

`.prob` is the probability that a trial came from the decision process, and
`.keep` is a policy applied to it. For every rule above the two coincide
because the probability is 0 or 1. For the mixture they come apart, and
`rt_screen()` makes the policy explicit. The default keeps a trial when
`.prob` exceeds `threshold`; raising the threshold removes more:

```{r policy}
p1_easy <- rt_example |>
  filter(id == "p1", condition == "easy")
mixture <- rule_mixture("lognormal")

at_50 <- rt_screen(p1_easy$rt, mixture)
at_90 <- rt_screen(p1_easy$rt, mixture, threshold = 0.9)
c(dropped_at_0.5 = sum(!at_50$.keep), dropped_at_0.9 = sum(!at_90$.keep))

table(keep = at_90$.keep, reason = at_90$.reason, useNA = "ifany")
```

The cross-table shows a distinction that matters when the exclusions are
reported. `.reason` records why the *rule* flagged a trial, and the mixture
flags a trial as a contaminant when the posterior is against it. The trials
the raised threshold removed on top of those were not flagged by the rule;
the policy dropped them, and their reason stays `NA`.

The other policy keeps each trial with probability `.prob`, drawing once per
trial. It is stochastic by design, and the seed belongs to the caller: there
is no `set.seed()` anywhere in the package.

```{r probabilistic}
set.seed(2026092)
drawn <- rt_screen(p1_easy$rt, mixture, policy = "probabilistic")
table(keep = drawn$.keep, reason = drawn$.reason, useNA = "ifany")
```

The same split appears: trials the rule flagged, and trials a draw against a
fractional `.prob` happened to remove. A probabilistic screen is the honest
way to carry the mixture's uncertainty into a resampling analysis, and the
aggregation article shows the alternative, which is to hand `.prob` to
`rt_summary()` as a weight and remove nothing.

## A rule that cannot be evaluated removes nothing

A spread of zero, a group of fewer than two trials, or a group below the
smallest size in van Selst and Jolicoeur's table leaves the rule without a
criterion. In every such case the rule keeps everything and says so in the
fits table, rather than returning `NA` or flagging by accident:

```{r unevaluable}
rt_screen(c(0.40, 0.40, 0.40), rule_sd(2.5))
screen_fits(c(0.40, 0.50, 0.60), rule_recursive("moving"))
```

`bmm::flag_contaminant_rts()` returns `NA` where its fit fails; `rtprep`
keeps the trial. A screen that cannot be evaluated has no grounds to remove
data, and a summary computed downstream should see every trial rather than
a hole.

## Every bound on one cell

The fits tables make the rules comparable in one picture. Each panel is the
same participant's hard condition; the dashed lines are the bounds that rule
reported, and the panel label carries the number of trials it removed.

```{r bounds-figure, fig.alt = "Six histograms of one participant's response times, one per screening rule, each with the rule's lower and upper bound drawn as dashed vertical lines. The absolute cutoff's bounds lie outside the data; the SD and recursive criteria show only an upper bound within the plotting range; the MAD criterion shows both; the EWMA chart shows a single lower cutoff."}
rules <- list(
  cutoff = rule_cutoff(0.18, 3),
  sd = rule_sd(2.5),
  mad = rule_mad(2.5),
  moving = rule_recursive("moving"),
  modified = rule_recursive("modified"),
  ewma = rule_ewma()
)

bounds <- bind_rows(lapply(names(rules), function(name) {
  fit <- screen_fits(p3_hard$rt, rules[[name]], response = p3_hard$response)
  data.frame(
    rule = name,
    lower = if (is.null(fit$lower)) fit$cutoff_rt else fit$lower,
    upper = if (is.null(fit$upper)) NA_real_ else fit$upper,
    n_dropped = fit$n_dropped
  )
}))
bounds

bounds <- bounds |>
  mutate(
    panel = paste0(rule, ": ", n_dropped, " removed"),
    panel = factor(panel, levels = panel)
  )

tidyr::expand_grid(p3_hard, bounds) |>
  ggplot(aes(rt)) +
  geom_histogram(binwidth = 0.05, fill = okabe_ito[1], alpha = 0.7) +
  geom_vline(
    data = bounds,
    aes(xintercept = lower), colour = okabe_ito[4], linetype = 2
  ) +
  geom_vline(
    data = filter(bounds, !is.na(upper)),
    aes(xintercept = upper), colour = okabe_ito[4], linetype = 2
  ) +
  coord_cartesian(xlim = c(0, 3)) +
  facet_wrap(~panel, ncol = 2) +
  labs(x = "Response time (s)", y = "Trials") +
  theme_minimal()
```

The absolute cutoff's bounds sit outside everything the task produced. The
SD criterion and both recursive criteria show only an upper bound, because
their lower bounds are negative and off the axis. The MAD criterion is the
one location-and-spread rule whose lower bound is positive, and it still
falls short of the fastest trial in the cell; in this cell the chart is the
only rule that removed anything from below. Which of those removals
were contaminants is a question the fits table cannot answer; the vignette
scores it against the generator's truth, and the ground-truth article does
so for a task matched to your own.

## Where next

`?rules` documents every constructor with the reference it implements and
the columns it adds to the fits table. Two further rules that the companion
simulation evaluated and did not recommend are kept unexported and described
in `?rules_experimental`. The mixture has
[its own article](mixture-screening.html), the aggregation article covers
what happens to the surviving trials, and `screen_compare()`
applies a whole roster at once and reports where the rules disagree.
